What to simulate? Inferring the right direction for mental rotation

### 2014-02-08 man/cogsci-2014/cogsci-2014.tex
Make model descriptions a little clearer

\begin{abstract}
  When people use mental imagery, how do they decide \textit{which}
  images to generate? To answer this question, we explored how mental
  simulation should be used in the classic psychological task of
  determining if two images depict the same object in different
  orientations \cite{Shepard1971}. Through a rational analysis of
  mental rotation, we formalized four models and compared them to
  human performance. We found that three models based on previous
  hypotheses in the literature were unable to account for several
  aspects of human behavior. The fourth, which is a model based on the
  idea of \textit{active sampling} \cite<e.g.,>{Gureckis:2012gu},
  provides plausible explanations where the other models do not. Based
  on these results, we suggest that the question of ``what to
  simulate?'' is more difficult than has previously been assumed, and
  that an active learning approach holds promise for uncovering the
  answer.

@@

Active sampling is the idea that when initial information is
incomplete, people will gather new information in a manner that
increases certainty about the problem space. This notion has recently
gained support in several areas of cognitive science
\cite<e.g.,>{Gureckis:2012gu}, including other spatial domains
\cite{Juni:2011vo}. In the case of mental rotation, if people do not
know the minimum angle beforehand, then actively performing rotations
may be the best way to gather evidence about the similarity between
the observed shapes.

@@

What if the point correspondences are unknown? A model which obeys
this constraint could first rotate in the direction that increases
local similarity, and then stop rotating once a ``match'' is found, as
determined by a threshold on the similarity $S$. If all rotations have
been exhausted, then flip, and try rotating again. We assume that the
locations where $S$ is greater than the threshold correspond to the
true $\theta$ (or points near the true $\theta$). So, as with the
\Oc{} model, we use Equation~\ref{eq:mle-lh-ratio}.

@@

In the current formulation of the problem, choosing the threshold is
straightforward because we know both the exact geometry of the shapes
and that a linear transformation exists which will perfectly align
them. However, this choice is not always clear \textit{a priori}, as
the global optimum depends on many factors (e.g., shape complexity,
dimensionality, perceptual uncertainty, and whether the shapes are
identical).  One way to deal with the problem of choosing a threshold
would simply be to perform an exhaustive search or use a global
optimization strategy. However, these strategies would not result in
the linear RT found by \citeA{Shepard1971}. A second alternative might
instead just perform a Hill Climbing (\Hc{}) search for a
\textit{local} maximum. Thus, as with the \Oc{} and \Th{} models, we
use Equation~\ref{eq:mle-lh-ratio}.

@@

Out of the 247 participants, 200 (81\%) were included in our
analyses. Of the other 47, we excluded 10 (4\%) because of an
experimental error, 6 (2.4\%) because they had already completed a
related experiment, and 31 (12.6\%) because they failed a
comprehension check. Comprehension was defined as correctly answering
at least 85\% of stimuli with a minimum rotation of $0^\circ$ or
$20^\circ$.  Additionally, we excluded 82 trials for which the RT was
either less than 100ms or greater than 20s.

@@

The \Hc{} was the only model for which there was no monotonic
relationship between rotation and RT
(Figure~\ref{fig:response-time-accuracy}). Moreover, the \Hc{} model
was barely above chance (\HcAccuracy{}) and there were \HcNumChance{}
individual stimuli for which it was not above chance. The \Hc{} model
was not a particularly good predictor of human RTs (\HcTimeCorr{}), as
shown in Figure~\ref{fig:human-model-scatters}. It was a moderately
good predictor of human accuracy (\HcAccuracyCorr{}).

@@

The \Oc{} and \Th{} models were---in terms of correlations---the best
predictors of human RTs. However, both are somewhat unsatisfying
explanations because they rely on \textit{a priori} knowledge that
people are unlikely to have. Moreover, they offer no explanation of
several aspects of human behavior. First, their overall RT
distributions look nothing like people's
(Figure~\ref{fig:histograms}).  Second, they both are 100\% accurate,
and so cannot explain the systematic relationship between rotation and
human accuracy (Figure~\ref{fig:response-time-accuracy}). Third,
neither model can explain the difference in people's behavior on
``same'' and ``flipped'' stimuli.

>>>>

\begin{abstract}
  When people use mental imagery, how do they decide \textit{which}
  images to generate? To answer this question, we explored how mental
  simulation should be used in the classic psychological task of
  determining if two images depict the same object in different
  orientations \cite{Shepard1971}. Through a rational analysis of
  mental rotation, we formalized four models and compared them to
  human performance. We found that three models based on previous
  hypotheses in the literature were unable to account for several
  aspects of human behavior. The fourth, which is based on the idea of
  \textit{active sampling} \cite<e.g.,>{Gureckis:2012gu}, provides
  plausible explanations where the other models do not. Based on these
  results, we suggest that the question of ``what to simulate?'' is
  more difficult than has previously been assumed, and that an active
  learning approach holds promise for uncovering the answer.

@@

Active sampling is the idea people gather new information in a manner
that increases certainty about the problem space. This notion has
recently gained support in several areas of cognitive science
\cite<e.g.,>{Gureckis:2012gu}, including other spatial domains
\cite{Juni:2011vo}. In the case of mental rotation, if people do not
know the minimum angle beforehand, then actively performing rotations
may be the best way to gather evidence about the similarity between
the observed shapes.

@@

A model which does not know point correspondences could use the
following algorithm: (1) pick a random direction; (2) take a single
step; (3) if that step decreased similarity, then begin rotating in
the reverse direction, otherwise continue rotating in the original
direction; (4) continue rotating in the chosen direction until a
``match'' is found (defined as finding a value of $S$ that exceeds a
threshold); and (5) if no match was found, flip, and start over from
step one.  We assume that the locations where $S$ is greater than the
threshold correspond to the true $\theta$ (or points near the true
$\theta$). So, as with the \Oc{} model, we use
Equation~\ref{eq:mle-lh-ratio}.

@@

In the current formulation of the problem, choosing the threshold is
straightforward because we know both the exact geometry of the shapes
and that a linear transformation exists which will perfectly align
them. However, this choice is not always clear \textit{a priori}, as
the global optimum depends on many factors (e.g., shape complexity,
dimensionality, perceptual uncertainty, and whether the shapes are
identical).  One way to deal with the problem of choosing a threshold
would simply be to perform an exhaustive search or use a global
optimization strategy. However, these strategies would not result in
the linear RT found by \citeA{Shepard1971}. A second alternative is to
perform a Hill Climbing (\Hc{}) search; i.e., rotate in the direction
that increases similarity until no further improvement can be
found. In contrast with the \Th{} model, this results in arriving in a
\textit{local} maximum (which may or may not be the global
maximum). Thus, as with the \Oc{} and \Th{} models, we use
Equation~\ref{eq:mle-lh-ratio}.

@@

Of the 247 participants, 200 (81\%) were included in our analyses. Of
the other 47, we excluded 10 (4\%) because of an experimental error, 6
(2.4\%) because they had already completed a related experiment, and
31 (12.6\%) because they failed a comprehension check, which was
defined as correctly answering at least 85\% of stimuli with a minimum
rotation of $0^\circ$ or $20^\circ$.  Additionally, we excluded 82
trials for which the RT was either less than 100ms or greater than
20s.

@@

The \Hc{} was the only model for which there was no monotonic
relationship between rotation and RT
(Figure~\ref{fig:response-time-accuracy}). Moreover, the \Hc{} model
was barely above chance (\HcAccuracy{}) and there were \HcNumChance{}
stimuli for which it was not above chance. The \Hc{} model was not a
good predictor of human RTs (\HcTimeCorr{}), as shown in
Figure~\ref{fig:human-model-scatters}. It was a moderately good
predictor of human accuracy (\HcAccuracyCorr{}).

@@

The \Oc{} and \Th{} models were the best predictors of human
RTs. However, both are somewhat unsatisfying explanations because they
rely on \textit{a priori} knowledge that people are unlikely to
have. Moreover, they offer no explanation of several aspects of human
behavior. First, their overall RT distributions look nothing like
people's (Figure~\ref{fig:histograms}).  Second, they both are 100\%
accurate, and so cannot explain the systematic relationship between
rotation and human accuracy
(Figure~\ref{fig:response-time-accuracy}). Third, neither model can
explain the difference in people's behavior on ``same'' and
``flipped'' stimuli.
